---
title: "Assignment 4: Exploratory Data Analysis"
author: "Caide Filipo"
date: "`r Sys.Date(9/18/26)`"
format: html
editor: visual
---

## Overview

Due: **Friday 9/25 11:59pm**

In this assignment, you will practice Exploratory Data Analysis (EDA) by making plots from a **Kaggle dataset of your choice**.

You will submit **at least 4 different plot types**, with short written interpretations for each.

Part 1 is worth a total of 1 point, and Part 2 is worth 4 points (1 point for each plot, question, and interpretation).

## Part 1: Get Your Data

1.  Go to [Kaggle Datasets](https://www.kaggle.com/datasets), create an account, and choose a dataset that interests you. Ensure your data has:
    1.  At least 4 variables
    2.  At least one categorical and one continuous variable
    3.  I recommend data that contains at least 50-100 observations
    4.  Tip: Open the CSV on Kaggle first and skim the columns to check that it meets these requirements
2.  Download the dataset as a CSV file.
3.  Place the CSV in your class R project `data` folder.
4.  Read it into R with `read_csv()`.

```{r, message=FALSE}
library(tidyverse)

mydata <- read_csv("data/student_exam_performance.csv")

```

### Describe your data

Data author:

Link:

Description:

### Preview your data

Use some of the strategies you've learned in this class or picked up elsewhere to summarize and quickly explore your data.

```{r}
view(mydata)
summary(mydata)
```

### Summary stats

Before you make plots, you must also:

- Show basic summaries of your dataset, including:

  - `count()` for categorical variables (to see category frequencies)

  - `mean()`, `median()`, `sd()`, `min()`, `max()` for at least one numeric variable

  - Remember to use `na.rm = TRUE` if you have missing values that affect summaries!

```{r}
install.packages("skimr")

count <- mydata
count <- count %>%
  count(class_participation)

count


summarise <- mydata
summarise <- summarise %>% 
  summarise(
    mean   = mean(previous_exam_score, na.rm = TRUE),
    median = median(previous_exam_score, na.rm = TRUE),
    sd     = sd(previous_exam_score, na.rm = TRUE),
    min    = min(previous_exam_score, na.rm = TRUE),
    max    = max(previous_exam_score, na.rm = TRUE)
  )
view(summarise)


```

Write 2–3 sentences here: What do the category counts tell you? What does the numeric summary show about spread, center, or outliers?

The category counts tell me that a graph of the participation level would have a "mountain" instead of a "valley". The numerical data shows that the minimum possible grade is a 30 and that the maximum would be a 100. A graph of the data would be nearly symmetrical as the mean and median are almost identical and that the main curve would be at \~70.

## Part 2: Visualize Your Data

Create **at least 4 plots**, each of a different type. Your plots may share variables, but each one should answer a **different question** about your data. At least one plot must include a variable that isn't in the others.

Each plot should have:

- A title and labeled axes

- A short interpretation (2-3 sentences) underneath

  - Sample interpretation: (boxplot of `price` by `cut` from `diamonds` data): *Ideal diamonds have the lowest median price, which is surprising since Ideal is the best cut. Every cut also has a lot of very expensive outliers. One possible reason for the low Ideal prices is that Ideal diamonds might be smaller on average, which I could check by plotting carat by cut.*

Use the EDA Part 3 notes as your reference. Since this is an EDA assignment, your plots can look plain, but consider adding colors, themes, or interactivity for interpretability and visual appeal if you know how.

### Plot 1

```{r}
# insert code here
ggplot(mydata, aes(x = previous_exam_score)) +
  geom_histogram(binwidth = 1) +
  theme_bw() +
  labs(title = "Exam Scores",
       subtitle = "Undergraduate and HS level performance on exams",
       x = "Previous Exam Score")
```

Question this plot answers: How were students doing before implementing studying techniques, preps, and limiting distractions?

Interpretation: Based on the resulting graph, the general body of the graph is centered on the 70 mark meaning a majority of students are passing. The level of higher achievers and lower achievers is around the same with a noticeable bump in those passing with a score of a 100.

### Plot 2

```{r}
# insert code here
ggplot(mydata, aes(x = study_hours_per_day, y = exam_score)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  theme_bw() +
  labs(
    title = "Study Hours vs. Final Exam Score",
    x = "Study Hours per Day",
    y = "Exam Score"
  )

```

Question this plot answers: What effect does how many hours studied have on the final exam score?

Interpretation: Generally, the more hours you study per day, the higher your exam score would be. But, any studying after an hour seems to have a higher chance of scoring a 100.

### Plot 3

```{r}
# insert code here

mydata$performance_level <- factor(
  mydata$performance_level, 
  levels = c("Low", "Medium", "High")
)


mydata %>% 
  drop_na(performance_level, sleep_quality) %>% 
  ggplot(aes(x = performance_level, y = fct_rev(sleep_quality))) +
  geom_count( color = "purple") +
  theme_bw() +
  labs(
    title = "Performance Level vs. Sleep Quality",
    x = "Performance Level",
    y = "Sleep Quality"
  )
```

Question this plot answers: What affect does the quality of your sleep have on the performance during an exam?

Interpretation: Based on the graph, majority of students taking the exam has poor to mediocre quality of sleep. Though, those with higher qualities of sleep tend to perform better on the exam.

### Plot 4

```{r}
# insert code here
ggplot(mydata, aes(x = study_method, y = exam_score, fill = study_method)) +
  geom_violin(alpha = 0.5, show.legend = FALSE) +
  geom_boxplot(width = 0.1, color = "black", alpha = 0.7, show.legend = FALSE) +
  labs(
    title = "Exam Score Density and Spread by Study Habit Category",
    x = "Study Habit Category",
    y = "Exam Score"
  )
```

Question this plot answers: Which study method is most reliable to get good grades?

Interpretation: These graphs for the different study methods look hard to distinguish. Doing practice tests seem to be the most predicting of a passing score as the distribution is narrowest.

------------------------------------------------------------------------

# Push to GitHub AND submit link on Canvas!!

Assignment is due Friday 9/25 at 11:59 pm.
