---
title: "CS-201 First Look at R"
author: "Amber Camp"
date: 8/26/26
format: html
editor: visual
toc: true
---

This is the first class session walking through code together!

We will be programming in R and in this class will primarily use `qmd` documents to keep things tidy and presentable. Each code chunk below is where you will run R code. Feel free to add notes as we go along.

## Navigating RStudio

By default, everyone should have the same four-pane layout:

1.  **Top left**: Source editor

2.  **Bottom left**: Console (and Terminal and Background Jobs)

3.  **Top Right**: Environment

4.  **Bottom Right**: Files, Plots, Help, etc.

### Customizing look & feel

Feel free to customize your RStudio.\
Go to Tools \> Global Options \> Appearance to select a color scheme you like.

I also personally like to enable these:

- Tools \> Global Options \> Code \> Insert matching parens/quotes

- Tools \> Global Options \> Console \> Show syntax highlighting in console input

I don't recommend rearranging your pane layout at this time.

## Running Code

Try running some code

### Load Library

```{r}
library (ggplot2)

```

### Some Scripting

```{r}
quiz1 <- 88 # this is how to assign a value to a name
quiz2 <- 94
quiz3 <- 79

quiz1

# averaging
(quiz1 + quiz2 + quiz3) / 3

mean(quiz1 + quiz2 + quiz3) # not correct, adds everything up
mean(quiz1, quiz2, quiz3) # wrong, only returns first arg
mean(c(quiz1, quiz2, quiz3))

list <- c(quiz1, quiz2, quiz3)

mean(list)

average <- mean(list)


```

```{r}
scores <- c(88,94,79,91,85)
scores

mean(scores)
min(scores)
max(scores)
length(scores)

summary(scores)

students <- c("a", "b", "c", "d", "e")
students

# bar plot
barplot(scores, names.arg = students)

barplot(scores, names.arg = students, ylim = c(0,100), main = "TITLE",
xlab = "pupils", y = "what they got")

```

## Introducing `ggplot2`

#### Make a dataframe

```{r}

grades_df <- data.frame(student = students, score = scores)
grades_df
summary(grades_df)

nrow(grades_df) # num of rows
ncol(grades_df) # num of cols
dim(grades_df) # df dimensions
```

#### Try ggplot

```{r}

ggplot(data = grades_df, aes(x = students, y = score)) # just the foundation


ggplot(data = grades_df, aes(x = reorder(students, -score), y = score, group = 1)) +
  geom_col(fill = "steelblue") +
  geom_point() +
  geom_line() +
  theme_minimal() +
  labs(title = "OLALALAL",
       subtitle = "walllalala",
       x = "bombahs",
       y = "cool stuff")


```

#### Mutate the dataframe

```{r}
library(tidyverse)

grades_df <- grades_df |> 
  mutate(study_hours = c(2,5,1,4,3))

```

#### Make a scatterplot

```{r}

ggplot(grades_df, aes(x = study_hours, y = score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_col(fill = "purple") +
  geom_line() +
  theme_dark() +
  labs(title = "WEHA",
       subtitle = "Blueberry",
       x = "SPONGAH",
       y = "SAD:C")

# cannot use fill (use color instead), don't use reorder



```

## Extra Credit (1pt)

What result do you expect when you run this expression?

```{r}

1 + 1 + 1

quiz1 + quiz2

1 + 1 + 1 - 3

0.1 + 0.1 + 0.1 - 0.3

options(scipen. = 999)

print(0.1, 20)
print(0.2, 20)
print(0.5, 20)
```

What did you observe?

> Answer:

Why do you think this is happening? (You can look online for help)

> Answer:
